---
title: "Prosper_Loan_Data_Udacity"
author: "Jayashree"
date: "4 May 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE, warning=FALSE)
```

## Prosper Loan Data
Load the csv file in to working directory
```{r}
getwd()
setwd('C:/Users/JAYASHREE/Desktop/Data Analyst Nanodegree/EDA/Project')
loan <- read.csv('prosperLoanData.csv',na.strings = c("","NA"))
summary(loan)
```

## Univariate Analysis

### Loan Amount
Let us analyse the loan amount taken by customers.We will plot the histogram to 
look at the distribution.
```{r}
library(ggplot2)
qplot(LoanOriginalAmount,data=loan,binwidth=700,color=I('black'),fill=I('#F79420'))
```


We could see from the plot that the distribution is positively skewed.Let us see
the summary statistics of the loan amount.

```{r}
summary(loan$LoanOriginal)
```

Let us take log transformation of Loan Amount and see the distribution
```{r}
qplot(LoanOriginalAmount,data=loan,color=I('black'),fill=I('#F79420'))+scale_x_log10()
```

The plot shows that the count of loan amount is bimodal peaking at around 7000 
and again at around 11000.

### Term

Let us take a look at the term column.We will plot a bar graph to understand the
distribution of the variable.
```{r}
length(unique(loan$Term))
ggplot(aes(factor(Term)),data=loan)+geom_bar(color=I('blue'),fill=I('red'),stat='count')+
  xlab('Term')
```


```{r}
table(loan$Term)
```

### Income Range
Let us look at the distribution of income range of borrowers.

```{r}
ggplot(aes(IncomeRange),data=loan)+geom_bar(color='blue',fill='pink')+
  theme(axis.text.x = element_text(size  = 10,
                            angle = 45,
                            hjust = 1,
                            vjust = 1))
```

From the plot we could see that income range are not arranged in ascending 
order.Let us make the income range as ordered variable.

```{r}
loan$IncomeRange <- ordered(loan$IncomeRange,levels=c("$0","$1-24,999","$25,000-49,999","$50,000-74,999",
                                                      "$75,000-99,999","$100,000+","Not displayed","Not employed"))
ggplot(aes(IncomeRange),data=loan)+geom_bar(color='blue',fill='violet')+
  theme(axis.text.x =  element_text(size  = 10,
                            angle = 45,
                            hjust = 1,
                            vjust = 1))
```

Most of the borrowers are in $25000-49999 Income range,closely followed by 
50,000-74999 range.

### Listing Category
Let us find out the purpose for which the borrowers have taken a loan.
```{r}
table(loan$ListingCategory)
```

We could see that for Category 1-Debt Consolidation,many borrowers have taken 
loan.

### Employment status
Let us look at the employment status of the borrowers.
```{r}
ggplot(aes(EmploymentStatus),data=subset(loan,!is.na(EmploymentStatus)))+geom_bar(color='black',fill='green')
```

Let us take log transformation of count so that lower values are visible

```{r}
ggplot(aes(EmploymentStatus),data=subset(loan,!is.na(EmploymentStatus)))+geom_bar(color='black',fill='green')+
  scale_y_log10()

```

### Unique borrowers in the dataset
Let us find how many unique borrowers are in the dataset.
```{r}
length(unique(loan$MemberKey))

```

Let us compute the percentage of first time borrowers
```{r}
length(unique(loan$MemberKey))/length(loan$MemberKey)
```

79.7 % of borrowers in the dataset are first time borrowers.

### LoanOriginationQuarter
Let us look at the loan origination quarter for borrowers.

```{r}
originationQuart <- substr(loan$LoanOriginationQuarter,1,2)
```
Let us find out in which quarter loans have originated the most.

```{r}
ggplot(aes(x=originationQuart),data=loan)+geom_bar(color = 'black', fill = '#099DD9',stat='count')
```

As seen from the plot most loans have originated in the 4th quarter.

### LoanOriginationDate
Let us look at the ditribution of origination of loans across days of the 
month.
```{r}
originationDay <- format(as.Date(loan$LoanOriginationDate,format="%Y-%m-%d"), "%d")
```

Let us plot bar graph to find the distribution of origination of loan across 
days of the month.
```{r}
ggplot(aes(originationDay),data=loan)+geom_bar(color = 'black', fill = '#088AD9',stat='count')
```

Most loans have originated in 30th of every month.

### Borrower State
Let us take a look at the distribution of states of the borowers from where they
have originated.
```{r}
ggplot(aes(BorrowerState),data=loan)+geom_bar(stat='count')+
  theme(axis.text.x = element_text(size  = 10,
                            angle = 90,
                            hjust = 1,
                            vjust = 1))
```

From the plot we could see that state of California has highest number of
borrowers.

### CurrentlyInGroup
Let us check the proportion of borrowers who are currenty in a group.
```{r}
ggplot(aes(CurrentlyInGroup),data=loan)+geom_bar(color='black',fill='blue')
```

We could see from the plot that  less than 1/4th of the borrowers are currently
in a group.

### StatedMonthlyIncome

Let us look at the summary statistics of the monthly income

```{r}
summary(loan$StatedMonthlyIncome)
```
We could see a huge difference between value in 3rd quartile and maximum value.
This shows the presence of outliers.


Let us look at the distribution of statedMontlyIncome of the borrowers within 
3rd quartile.
```{r}
ggplot(aes(StatedMonthlyIncome),data=loan)+geom_histogram(binwidth=500,color='black',fill='purple')+
  scale_x_continuous(limits=c(0,7000))
```

From the plot we could see that distribution is nearly normal.

### IsBorrowerHomeOwner

Let us look at the distribution of home ownership of borrowers.
```{r}
table(loan$IsBorrowerHomeowner)
ggplot(aes(x=IsBorrowerHomeowner),data=loan)+geom_bar(color='red',fill='yellow')
```

Borrowers with home ownership are slightly higher than borrowers without home 
ownership.
## Bi-variate Analysis

### StatedMonthlyIncome Vs LoanOriginalAmount
Let us find if monthly income of the borrower has an impact on his loan amount.
```{r}
ggplot(aes(x=LoanOriginalAmount,y=StatedMonthlyIncome),data=loan)+geom_point()
```

In the above figure,data points are concentrated near the x-axis,so we shall 
limit the y-axis values to take a closer look at the data points.

```{r}
ggplot(aes(x=LoanOriginalAmount,y=StatedMonthlyIncome),data=loan)+geom_jitter(alpha=1/20,position='jitter')+
  scale_y_continuous(limits=c(0,50000))+geom_smooth(method=lm)
```


From the figure we could not find a linear relationship.Let us compute 
correlation coefficient and see what it tells about the strength between loan 
amount and monthly salary.

```{r}
with(data=loan,cor.test(LoanOriginalAmount,StatedMonthlyIncome))
```

### Employment Status vs Loan Amount
I assume that  Employed people are eligible for higher loan amount than other 
category of people.Let us plot the relation between these 2 variables to find 
whether my assumptions hold right.
```{r}
ggplot(aes(x=EmploymentStatus,y=LoanOriginalAmount),data=subset(loan,!is.na(EmploymentStatus)))+geom_boxplot()
```

My assumptions holds true that Employed persons get higher loan amount.
Let us look at the summary statistics for each employment group.
```{r}
by(loan$LoanOriginalAmount,loan$EmploymentStatus,summary)
```

### Let us the see the proportion of borrower home owner
```{r}
table(loan$IsBorrowerHomeowner)
```

I assume that the loan amount will be higher for the borrower with home.Let us 
find out if my assumption is right by plotting a frequency polygon.

```{r}
ggplot(aes(x = LoanOriginalAmount, y = ..count../sum(..count..)), data = loan) + 
  geom_freqpoly(aes(color =IsBorrowerHomeowner)) + 
  scale_x_continuous(breaks=seq(0,36000,5000))+
  xlab('Loan Amount') + 
  ylab('Percentage of borrowers with that loan amount')
```


From the plot we could see that for lower amounts of loan amount the percentage 
of borrowers without home ownership is higher but as the loan amount increases 
home owner borrower has higher percentage of loan amount.

### Employment status vs Amount delinquent
Let us find the distribution of amount delinquent for various categories of 
employment status.
```{r}
ggplot(aes(y=AmountDelinquent,x=EmploymentStatus),data=subset(loan,!is.na(EmploymentStatus)))+geom_bar(stat='identity')
```

From the plot we could infer that Employed borrowers has high amount delinquent.

### Employment status Duration vs Amount Delinquent
Let us look at the relationship of amount delinquent to the number of months 
the borrower was employed.
```{r}
ggplot(aes(x=EmploymentStatusDuration,y=AmountDelinquent),data=loan)+geom_point()
```

From the plot we could infer that the lesser the number of months employed,the 
higher is the amount delinquent.Since data points are concentrated at the bottom
lets change the scale for y axis to get clear view of data.

```{r}
ggplot(aes(x=EmploymentStatusDuration,y=AmountDelinquent),data=loan)+
  geom_jitter(alpha=1/6,position='jitter')+scale_y_continuous(limits=c(0,200000))+
  geom_smooth(method=lm)+scale_y_log10()
```

Let us find the correlation between these 2 variables
```{r}
with(loan,cor.test(EmploymentStatusDuration,AmountDelinquent))
```

We could see from the correlation coefficient that both the vairables are 
weakly related.

### InvestmentFromFriendsCount vs InvestmentFromFriendsAmount
I assume that as more friends chips in the amount collected will be high.Let us 
see if that assumption is correct.
```{r}
ggplot(aes(x=InvestmentFromFriendsCount,y=InvestmentFromFriendsAmount),data=loan)+geom_jitter()+
  geom_smooth(method=lm)
```

From the plot we could see that my assumption is wrong.There is no strong 
relationship between friends count and friends amount.

```{r}
table(loan$InvestmentFromFriendsCount)
```

Only few borrowers have friends chiped in to lend the loan.

```{r}
by(loan$InvestmentFromFriendsAmount,loan$InvestmentFromFriendsCount,summary)
```

###  Employment status vs Investors
Let us look at the distribution of investors who have lended the loan.

```{r}
ggplot(aes(x=EmploymentStatus,y=Investors),data=subset(loan,!is.na(EmploymentStatus)))+
  geom_bar(stat='identity',color=I('blue'))
```

As seen from the plot investors are funding high for employed borrowers.


## Multivariate Analysis

### Monthly Income vs Loan Amount based on Employment Status

Let us look at plot of monthly income vs loan amount sanctioned for each
category of employed persons.
```{r}
ggplot(aes(x=LoanOriginalAmount,y=StatedMonthlyIncome),data=subset(loan, !is.na(EmploymentStatus) ))+geom_point(color=I('#F79420'),alpha=1/10)+  scale_y_continuous(limits=c(0,50000))+facet_wrap(~EmploymentStatus)+
  geom_smooth(method=lm)
```

Full time Employed persons have higher salary and hence higher loan amount.

### CurrentDelinquencies vs loan amount for various categories of 
### Employmentstatus
Let us calculate median loan amount for borrowers who have current delinquencies
for varied category of employment status.

```{r}
ggplot(aes(x=CurrentDelinquencies,y=LoanOriginalAmount),data=subset(loan,!is.na(EmploymentStatus)))+
  geom_smooth(aes(color=EmploymentStatus),method ="lm", formula = y ~ x+I(x^2), size = 1 )+scale_color_brewer(type='qual')  
  
```

Employed borrowers have less number of delinquency compared to other category of
borrowers.

### Borrowers APR vs Borrower Rate based on Term

Let us find if term of loan impact the interest rate and Annual Percentage Rate

```{r}

ggplot(aes(x = BorrowerAPR,y = BorrowerRate  , color = factor(Term)), data = loan) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
      scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))
```

As term increases,borrower APR increases with borrower interest rate forming a 
linear relationship.Let us look at the correlation coefficient.

```{r}
with(loan,cor.test(BorrowerAPR,BorrowerRate))
```

This shows that both the variables have strong linear relationship.

### Income Range vs Stated Monthly Income based on Income verifiable

Let us look at the distribution of montly income of borrowers based on their 
income range and verificaton of income.

```{r}
ggplot(aes(x=IncomeRange,y=StatedMonthlyIncome, fill = IncomeVerifiable),data=loan)+
  geom_boxplot( )+scale_y_log10() +
  theme(axis.text.x =  element_text(size  = 10,
                            angle = 45,
                            hjust = 1,
                            vjust = 1))
```

The plot shows that only few borrowers have given information about verification
of income.

## Final Plots and Summaries

### Plot One

#### Employment Status vs Loan Original Amount

Loan Amount is highly influenced by employment status of the borrower.Full time 
employed borrowers earn higher loan amount compared to other categories of 
borrowers.

```{r}
ggplot(aes(x=EmploymentStatus,y=LoanOriginalAmount),data=subset(loan,!is.na(EmploymentStatus)))+geom_boxplot()+
  ylab('Loan Original Amount [USD]')+ggtitle('Employment Status vs Loan Original Amount')
```

The boxplot shows that the median loan amount for employed borrowers are higher.
Also lots of outliers are present for full-time and not available category of
borrowers.Let us look at the summary statistics for the plot.

```{r}
by(loan$LoanOriginalAmount,loan$EmploymentStatus,summary)

```

I have chose this plot as this shows clear trend.

### Plot Two

#### Percentage of home owners vs Loan Original Amount

For lower amounts of loan home ownership is not a major influencing factor.But 
for higher loan amount,home ownership becomes an important factor.This may be 
due to the reason that house document can be mortaged incase of failure to repay
the loan.

```{r}
ggplot(aes(x = LoanOriginalAmount, y = ..count../sum(..count..)), data = loan) + 
  geom_freqpoly(aes(color =IsBorrowerHomeowner)) + 
  scale_x_continuous(breaks=seq(0,36000,5000))+
  xlab('Loan Original Amount [USD]') + 
  ylab('Percentage of borrowers with that loan amount')+
  ggtitle('Percentage of Home Owners with respect to Loan Amount (%)')
```

Let us look at the proportion of home owners in the dataset.

```{r}
table(loan$IsBorrowerHomeowner)
```



### Plot Three

#### Borrowers APR vs Borrower Rate based on Term

As term increases there is a linear relationship between Borrower APR and 
Borrower Rate.

```{r}
ggplot(aes(x = BorrowerAPR,y = BorrowerRate  , color = factor(Term)), data = loan) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
      scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))+
  ggtitle('Borrower APR vs Borrower Rate based on Term')+
  xlab('Borrower APR (%)')+
  ylab('Borrower Rate (%)')

```

Let us calculate correlation coefficient to check for linearity.

```{r}
with(loan,cor.test(BorrowerAPR,BorrowerRate))
```

The values shows that both the variables have strong linear relationship.


## Reflection

Prosper Loan dataset is a huge dataset of more tha 1 lakh data.As I explored the
data,I found success in finding pattern of loan amount based on employment 
status and home ownership.I found difficult to choose the variables to explore
as there are 81 of them containing varied and vast information.I have analysed 
around 20 variables.There is a lot of scope in taking this project forward.
All the variables can be analysed and plotted to check for the trend and model
can be built.